\pkg{dataMaid}: your assistant for data documentation in \proglang{R}

### 2017-10-20 latex/article_vol3.tex
Small fixes in the article.

\newcommand{\hl}[1]{\textcolor{magenta}{#1}}

@@

%% almost as usual
\author{Anne Helby Petersen\\Biostatistics\\Department of Public
  Health\\University of Copenhagen \And Claus Thorn Ekstr\o m\\Biostatistics\\Department of Public
  Health\\University of Copenhagen}
\title{\pkg{dataMaid}: your assistant for data documentation in \proglang{R}}

@@

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Claus Thorn Ekstr\o m\\
  Biostatistics, Department of Public Health\\
  University of Copenhagen\\
  Denmark\\
  E-mail: \email{ekstrom@sund.ku.dk}\\
  URL: \url{http://staff.pubhealth.ku.dk/~ekstrom/}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

@@

But even when tools are available for identifying problems in a dataset, the activity of data cleaning still suffers from a challenge that has recieved increasing attention in the scientitic communities in the later years: Data cleaning is not very straight forward to document and therefore, reproducibility suffers. \pkg{dataMaid}'s most central purpose is to facilitate a data cleaning workflow where documentation is thoroughly integrated rather than a secondary add-on. This is accomplished by centering the data cleaning process around auto-generated data overview reports that summarize the contents of the current version of a dataset and flags potential problems that could be candidates for the next bullet on the cleaning agenda.

A number of \proglang{R} packages made for other pre-analysis tools are
already available, including \pkg{janitor} \citep{janitor}, \pkg{dplyr}
\citep{dplyr}, \pkg{tidyr}\citep{tidyr}, \pkg{data.table} \citep{data.table}, \pkg{DataCombine}
\citep{DataCombine} , \pkg{validate} \citep{validate}, \pkg{assertr} \citep{assertr}, and
\pkg{DataExplorer} \citep{DataExplorer}. These packages focus on
different stages of the pre-analysis work, most of which are not
really data cleaning.  \pkg{janitor} provides tools for data import
with a particular emphasis on the challenges of getting neat data
frames from Microsoft Excel data files. \pkg{dplyr}, \pkg{tidyr}, \pkg{data.table}
and \pkg{DataCombine} go a few steps further by providing a wide array
of extremely powerful tools for data wrangling, including a number of
particularly useful functions for merging and working with very large
datasets. When it comes to actual data cleaning, however, the options
are fewer.  \pkg{validate} (and the similar packages \pkg{editrules}
\citep{editrules} and \pkg{deducorrect} \citep{deducorrect} from the
same authors) does offer a few tools for identifying and
auto-correcting errors in a dataset, but, as the name suggests, the
focus is on internal validity rather than general data cleaning. In
practice, this means that quite elegant tools for, e.g., linear
restraints among the variables in a dataset can be applied, but
looking for potentially miscoded missing values is not really
feasible. The main difference between these two challenges is the
direction in which the data is inspected: While linear constraints
work observation-wise with no ambiguity, determining whether or not something
is a miscoded missing value often requires knowledge about the full
variable (e.g. range or data type), and thus it should be performed
variable-wise. \pkg{validate} does not allow for user-defined extensions
of the latter type, thereby limiting its data cleaning potential.
Automatic data correction functions are also provided by
\pkg{validate} which we consider to be quite a dangerous cocktail: all
power is given to the the computer with no human supervision, and
investigators are less likely to make an active, case-specific choice
regarding the handling of the potential errors.  Finally, no tools
have been made to easily document exactly which checks and preliminary
results were used in the data cleaning process. \hl{Something about \pkg{assertr} here.}

All in all, the large role of data cleaning in any data analysts
everyday endeavors is hardly matched in the amount of available
\proglang{R} software solutions. In particular, few packages attempt
to implement systematic, reproducible data cleaning. And while the
available tools attempt to alleviate the ubiquitous ad hoc approach to
data cleaning, they are primarily intended for the data savvy users
and less so for the general researcher with a knowledge about a
specific field and the context of the available data. The
\pkg{dataMaid} package \citep{dataMaid} presented here tries to address this by
providing a framework that both allows for extendable, systematic,
reproducible data cleaning, and summarizing findings for researchers
from other fields such that they can act as human experts when
tracking down potential errors.

@@

This manuscript is structured as follows: First, in Section
\ref{sec:usingdataMaid}, we introduce the main representative of the first
paradigm, namely the \code{makeDataReport()} function, which generates data
overview documents. In the \pkg{dataMaid} package, we have
provided a number of default checks that cover the data
cleaning challenges, we find to be most common and these are also summarized in Section \ref{sec:usingdataMaid}. Next, in Section
\ref{sec:interactiveCleanR}, we present the interactive mode of \pkg{dataMaid}, as motivated
by the second paradigm above. Next up, we show how the data report and the interactive mode of \pkg{dataMaid} can be combined step-by-step to conduct a well-documented, systematic data cleaning in Section \ref{sec:bigExample}. Here, we asses and clean a dirty dataset with information about the US presidents.  At last, in Section \ref{sec:specificExamples}, we discuss a number of examples of
specific data cleaning- and documentation challenges and how \pkg{dataMaid} can be used to
solve them.

@@

% \hl{Is there a Section 6?}

%\hl{Do they typically discuss notation in articles like this? I try (somewhat inconsistently) to refer to functions as \R{function()} while other \R{R} objects are referred to merely as \R{object}. This is also the style adopted by Wickham in his books.}

@@

The \code{makeDataReport()} function is the primary workhorse of \pkg{dataMaid} and
this is the only function needed if a standard battery
of tests are used to generate data reports. The data report itself is an overview document, intended for reading
by humans, in either pdf, html or word (.docx) format. Appendix \ref{sec:appendix1}
provides an example of a data report, produced by
calling \code{makeDataReport()} on the dataset \code{toyData} available in
\pkg{dataMaid}. The first two pages (excluding the frontpage) of this data report are
shown in Figure~\ref{fig:example1}. \code{toyData} is a very
small ($15$ observations of $6$ variables), artificial dataset which was created to
illustrate the main capabilities of \pkg{dataMaid}. The following
commands load the dataset and produce the report:

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=2]{dataMaid_toyData.pdf}}
\frame{\includegraphics[width=7.5cm,page=3]{dataMaid_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{dataMaid_testData.pdf}
\end{center}
\caption{The two first pages of the report created by running \code{makeDataReport()} on the \code{toyData}
  dataset. First a summary of the full dataset is given along with an overview of what checks were performed. Next, a summary of all the variables aand whether or not they are problematic is provided. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\label{fig:example1}
\end{figure}

@@

By default, an \proglang{R} markdown file and a rendered pdf \hl{OBS: måske html/doc alt efter hvem der kalder funktionen?} overview
document is produced, saved to the working directory and
opened for immediate inspection. Such a data report consists of three parts, two of which 
parts. First, an overview of what was done is presented under the
title \textit{Data report overview}. Secondly, an index listing each variable along with an indication of whether it was found to be problematic or not is provided. Thirdly, each variable in the
dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution and a
list of flagged issues, if any. For instance, in the \code{numeric}-type variable
\code{var2} from \code{toyData}, \code{makeDataReport()} has identified two values that
are suspected to be miscoded missing values (\code{999} and \code{NaN}),
while two values were also flagged as potential outliers that should
be investigated more carefully.

@@

\smallskip Control input variables and meta information\\
\quad \code{useVar} & What variables should be used? & \code{NULL} (corresponding to all variables) \\
\quad \code{ordering} & Ordering of the variables in the data summary (as is or alphabetical) & \code{"asIs"} \\
\quad \code{onlyProblematic} & Should only variables flagged as problematic be included in the \textit{Variable list}? & \code{FALSE} \\
\quad \code{listChecks} & Should an overview of what checks were performed be listed in the \textit{Data report overview}? &  \code{TRUE} \\
\quad \code{preChecks} & What check functions should be called to determine whether a variable is suitable for summarization, visualization and checking? & \code{c("isKey", "isSingular", "isSupported")} \\
\quad \code{reportTitle} & What should the title displayed on the front page of the report be? & \code{NULL} (corresponds to the dataset name) \\

\smallskip Control summarize, visualize, and check steps \\
\quad \code{mode} & What steps should be performed for each variable (out
                 of the three possibilities \textit{summarize},
                 \textit{visualize}, \textit{check})? &
                                                        \code{c("summarize", "visualize", "check")} (corresponding to all three steps) \\
\quad \code{smartNum} & Should numerical values with only a few unique
                     levels be flagged and treated as a factor variable? & \code{TRUE} \\
\quad \code{maxProbVals} & Maximum number of problematic values to print, if any are found in data checks & \code{10} \\
\quad \code{maxDecimals} & Maximum number of decimals to print for numeric values in the variable list & \code{2} \\
\quad \code{twoCol} & Should the summary table and visualizations be placed side-by-side (in two columns)? & \code{TRUE} \\

\smallskip Control output and post-processing \\
\quad \code{output} & Type of output file to be produced (html, word (.docx) or pdf) & \code{"pdf"} \hl{OBS. Ændrer vi dette?}\\
\quad \code{render} & Should the output file be rendered from markdown? & \code{TRUE} \\
\quad \code{openResult} & If a  pdf/html file is rendered, should it
                       automatically open afterwards, and if not,
                       should the \code{rmarkdown} file be opened? & \code{TRUE} \\
\quad \code{replace} & Overwrite an existing file with the same name? & \code{FALSE} \\

@@

\subsection{Polishing off the arguments}
We begin with an example that is intended as an illustration of how
\code{makeDataReport()} might be used in the very first stages of data
cleaning, when the we are uncertain about the complexities of the
errors and how much time should be allocated to data cleaning. At
this stage, what is really needed, is a very rough idea of the
severity of errors in the dataset. In this scenario, we might wish to
obtain a summary document in html format that only contains the
variables with potential problems, and with a limit of, say, maximum
2 printed potential errors for each variable. Also, we can add the
argument \code{replace=TRUE} in order to force \code{makeDataReport()} to
overwrite any existing files produced by \code{makeDataReport()}.  Using the
\code{toyData} dataset as a guinea pig, we type:

@@

The final rendering of the generated markdown file is controlled by
the \code{render} and \code{openResult} arguments, which both default to
\code{TRUE}. \code{render} determines if the \proglang{R} markdown file
produced should be rendered using the \pkg{rmarkdown} \citep{rmarkdown} package and
\code{openResult} decides whether the output html or pdf file should be
opened. The following command produces an \proglang{R} markdown file
containing the information needed for generating a data report, but without
rendering nor opening the markdown file:

@@

\begin{table}
\centering
\begin{tabular}{p{0.35\linewidth} p{0.3\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth}
 p{0.01\linewidth} p{0.01\linewidth}}
  \hline
& Description &  \multicolumn{7}{c}{Variable classes} \\ \smallskip
 & &  C & F & I & L & B & N & D\\
  \hline \smallskip
  \textbf{\code{summaryFunction}s}  \smallskip \\
  \quad \code{centralValue} & Compute median for numeric variables, mode for categorical variables &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \quad \code{countMissing} & Compute proportion of missing observations &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \quad \code{minMax} & Find minimum and maximum values &   &  & $\times$ & &  & $\times$ & $\times$  \\
  \quad \code{quartiles} & Compute 1st and 3rd quartiles &    &  & $\times$ & &  & $\times$ & $\times$ \\
  \quad \code{uniqueValue} & Count number of unique values &   $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \quad \code{variableType} & Data class of variable & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
  \smallskip \\
 \textbf{\code{visualFunction}s} \smallskip \\
  \quad \code{basicVisual} & Histograms and barplots using base \proglang{R} graphics &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \quad \code{standardVisual} & Histograms and barplots using \pkg{ggplot2} &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \smallskip \\
 \textbf{\code{checkFunction}s} \smallskip \\
 \quad \code{identifyCaseIssues} & Identify case issues &  $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyLoners} & Identify levels with $<$ 6 obs. & $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyMissing} & Identify miscoded missing values &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &  \\
 \quad \code{identifyNums} & Identify misclassified numeric or integer variables & $\times$ & $\times$ & & & & &  \\
 \quad \code{identifyOutliers} & Identify outliers &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyOutliersTBStyle} & Identify outliers (Turkish Boxplot style) &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyWhitespace} & Identify prefixed and suffixed whitespace &  $\times$ & $\times$ & & $\times$ & & &  \\
 \quad \code{isCPR} & Identify Danish CPR numbers & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &$\times$   \\
 \quad \code{isSingular} & Check if the variable contains only a single value & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$  \\
 \quad \code{isKey} & Check if the variable is a key & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$\smallskip   \\
 \quad \code{isSupported} & Check if the variable is among the supported variable types & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$\smallskip   \\
 \hline
\end{tabular}
\caption{List of all summary functions (used in the summary table for
  each variable in the output), visual functions (used for  visualization of each variable), and
  check functions (used for data checks for each variable) currently implemented in \pkg{dataMaid}. The variable
  classes C, F, I, L, B, N, and D, refer to \code{character}, \code{factor},
  \code{integer}, \code{labelled}, \code{logical} (boolean), \code{numeric}, and \code{Date}, respectively.}
\label{table.SVCfunctions}
\end{table}

We will now move on to discuss how not only the structure of the
data assesment steps is manipulated, but also its very contents. This is
done through the summarize/visualize/check (SVC) steps, as illustrated
in Figure \ref{figure:cleanStructure}.  \pkg{dataMaid} uses three
different types of functions for performing these steps, namely
\code{summaryFunction}s, \code{visualFunction}s and
\code{checkFunction}s.  By default, \code{makeDataReport()} runs selected summary,
visualization and check functions on each variable in the dataset, and
the exact choice of these functions depends on the classes of the
variables. For instance, though detection of outlier values might be
interesting for numerical variables, it holds little meaning for
factor variables, and therefore, numerical and factor variables need
different checks. Table~\ref{table.SVCfunctions} lists all available
summarize/visualize/check functions, but we can also use the
\code{allSummaryFunctions()}, \code{allVisualFunctions()}, and
\code{allCheckFunctions()} functions in \pkg{dataMaid} to print overview
lists in \proglang{R}. For example, the implemented
\code{summaryFunction}s are:

@@

We can control what summaries and checks are applied for each variable type
through the \code{summaries}, \code{visuals} and \code{checks} arguments. Each of these arguments should be a list with one entry for each variable type and a number of function names for each such entry. The easiest way to specify the arguments is by use of the built-in helper functions \code{setSummaries()}, \code{setVisuals()} and \code{setChecks()}.

For instance, we can inspect the default settings for summaries by calling

\begin{Schunk}
\begin{Sinput}
R> setSummaries()
\end{Sinput}
\begin{Soutput}
$character
[1] "identifyMissing"  "identifyWhitespace"  "identifyLoners"    
[4]  "identifyCaseIssues"  "identifyNums"

$factor
[1] "identifyMissing"  "identifyWhitespace"  "identifyLoners"     
[4] "identifyCaseIssues"  "identifyNums"

$labelled
[1] "identifyMissing"  "identifyWhitespace"  "identifyLoners"     
[4] "identifyCaseIssues"  "identifyNums"

$numeric
[1] "identifyMissing"  "identifyOutliers"

$integer
[1] "identifyMissing"  "identifyOutliers"

$logical
NULL

$Date
[1] "identifyOutliers"
\end{Soutput}
\end{Schunk} %$

>>>>

\newcommand{\hl}[1]{\textcolor{magenta}{#1}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}

@@

%% almost as usual
\author{Anne Helby Petersen\\Section of Biostatistics\\Department of Public
  Health\\University of Copenhagen \And Claus Thorn Ekstr\o m\\Section of Biostatistics\\Department of Public
  Health\\University of Copenhagen}
\title{\pkg{dataMaid}: your assistant for data documentation in \proglang{R}}

@@

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Claus Thorn Ekstr\o m\\
  Section of Biostatistics, Department of Public Health\\
  University of Copenhagen\\
  Denmark\\
  E-mail: \email{ekstrom@sund.ku.dk}\\
  URL: \url{http://staff.pubhealth.ku.dk/~ekstrom/}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

@@

But even when tools are available for identifying problems in a dataset, the activity of data cleaning still suffers from a challenge that has recieved increasing attention in the scientitic communities in the later years: Data cleaning is not very straight forward to document and therefore, reproducibility suffers. We present a new \proglang{R} package, \pkg{dataMaid} \citep{dataMaid}, whose most central purpose is to facilitate a data cleaning workflow where documentation is thoroughly integrated rather than a secondary add-on. This is accomplished by centering the data cleaning process around auto-generated data overview reports that summarize the contents of the current version of a dataset and flags potential problems that could be candidates for the next bullet on the cleaning agenda.

A number of \proglang{R} packages made for other pre-analysis steps are
already available, including \pkg{janitor} \citep{janitor}, \hl{assertive}, \pkg{dplyr}
\citep{dplyr}, \pkg{tidyr}\citep{tidyr}, \pkg{data.table} \citep{data.table}, \pkg{DataCombine}
\citep{DataCombine} , \pkg{validate} \citep{validate}, \pkg{assertr} \citep{assertr}, and
\pkg{DataExplorer} \citep{DataExplorer}. These packages focus on
different stages of the pre-analysis work, most of which are not
really data cleaning.  \pkg{janitor} provides tools for data import
with a particular emphasis on the challenges of getting neat data
frames from Microsoft Excel data files. \hl{Mention assertive here}. \pkg{dplyr}, \pkg{tidyr}, \pkg{data.table}
and \pkg{DataCombine} go a few steps further by providing a wide array
of extremely powerful tools for data wrangling, including a number of
particularly useful functions for merging and working with very large
datasets. When it comes to actual data cleaning, however, the options
are fewer.  \pkg{validate} (and the similar packages \pkg{editrules}
\citep{editrules} and \pkg{deducorrect} \citep{deducorrect} from the
same authors) does offer a few tools for identifying and
auto-correcting errors in a dataset, but, as the name suggests, the
focus is on internal validity rather than general data cleaning. In
practice, this means that quite elegant tools for, e.g., linear
restraints among the variables in a dataset can be applied, but
looking for potentially miscoded missing values is not really
feasible. The main difference between these two challenges is the
direction in which the data is inspected: While linear constraints
work observation-wise with no ambiguity, determining whether or not something
is a miscoded missing value often requires knowledge about the full
variable (e.g. range or data type), and thus it should be performed
variable-wise. \pkg{validate} does not allow for user-defined extensions
of the latter type, thereby limiting its data cleaning potential.
Automatic data correction functions are also provided by
\pkg{validate} which we consider to be quite a dangerous cocktail: all
power is given to the the computer with no human supervision, and
investigators are less likely to make an active, case-specific choice
regarding the handling of the potential errors.  Finally, no tools
have been made to easily document exactly which checks and preliminary
results were used in the data cleaning process. The \pkg{assertr} package provide very similar - and very nice - tools to those of \pkg{validate}, but without any amibitions of conducting auto-cleaning.

%All in all, the large role of data cleaning in any data analysts
%everyday endeavors is hardly matched in the amount of available
%\proglang{R} software solutions. In particular, few packages attempt
%to implement systematic, reproducible data cleaning. And while the
%available tools attempt to alleviate the ubiquitous ad hoc approach to
%data cleaning, they are primarily intended for the data savvy users
%and less so for the general researcher with a knowledge about a
%specific field and the context of the available data. The
%\pkg{dataMaid} package \citep{dataMaid} presented here tries to address this by
%providing a framework that both allows for extendable, systematic,
%reproducible data cleaning, and summarizing findings for researchers
%from other fields such that they can act as human experts when
%tracking down potential errors.
%Jeg har slettet dette: Synes det dækkes ok af det nye introafsnit før præsentationen af andre pakker og så er introduktionen ikke blevet længere alt i alt.

@@

This manuscript is structured as follows: First, in Section
\ref{sec:usingdataMaid}, we introduce the main representative of the first
paradigm, namely the \code{makeDataReport()} function, which generates data
overview documents. In the \pkg{dataMaid} package, we have
provided a number of default checks that cover the data
cleaning challenges we find to be most common and these are also summarized in Section \ref{sec:usingdataMaid}. Next, in Section
\ref{sec:interactiveCleanR}, we present the interactive mode of \pkg{dataMaid}, as motivated
by the second paradigm above. Next, we show atep-by-step how the data report and the interactive mode of \pkg{dataMaid} can be combined to conduct a well-documented, systematic data cleaning in Section \ref{sec:bigExample}. Here, we assess and clean a dirty dataset with information about the US presidents.  At last, in Section \ref{sec:specificExamples}, we discuss a number of examples of
specific data cleaning- and documentation challenges and how \pkg{dataMaid} can be used to
solve them.

@@

@@

The \code{makeDataReport()} function is the primary workhorse of \pkg{dataMaid} and
this is the only function needed if a standard battery
of tests are used to generate data reports. The data report itself is an overview document, intended for reading
by humans, in either pdf, html or word (.docx) format. Appendix \ref{sec:appendix1}
provides an example of a data report, produced by
calling \code{makeDataReport()} on the dataset \code{toyData} available in
\pkg{dataMaid}. The first two pages (excluding the frontpage) of this data report are
shown in Figure~\ref{fig:example1} and the following two pages are shown in Figure~\ref{fig:example2}. \code{toyData} is a very
small ($15$ observations of $6$ variables), artificial dataset which was created to
illustrate the main capabilities of \pkg{dataMaid}. The following
commands load the dataset and produce the report:

\begin{figure}[tb]
\begin{center}
\frame{\includegraphics[width=7.5cm,page=2]{dataMaid_toyData.pdf}}
\frame{\includegraphics[width=7.5cm,page=3]{dataMaid_toyData.pdf}}
%\includepdf[pages={2}, pagecommand={}]{dataMaid_testData.pdf}
\end{center}
\caption{The two first pages of the report created by running \code{makeDataReport()} on the \code{toyData}
  dataset. First, a summary of the full dataset is given along with an overview of what checks were performed. Next, a summary of all the variables and whether or not they are problematic is provided. Larger versions of the pages can be seen in
  Appendix~\ref{sec:appendix1}.}
\label{fig:example1}
\end{figure}

@@

By default, an \proglang{R} markdown file and a rendered pdf \hl{OBS: måske html/doc alt efter hvem der kalder funktionen?} overview
document is produced, saved to the working directory and
opened for immediate inspection. Such a data report consists of three parts, two of which 
are presented in Figure~\ref{fig:example1}. First, an overview of what was done is presented under the
title \textit{Data report overview}. Secondly, an index listing each variable along with an indication of whether it was found to be problematic or not is provided. Thirdly, as seen in Figure~\ref{fig:example2}, each variable in the
dataset is presented in turn using (up to) three tools in the
\textit{Variable list}: A table summarizing key features of the
variable, a figure visualizing its distribution and a
list of flagged issues, if any. For instance, as shown in Figure~\ref{fig:example2}, for the \code{numeric}-type variable
\code{var2} from \code{toyData}, \code{makeDataReport()} has identified two values that
are suspected to be miscoded missing values (\code{999} and \code{NaN}),
while two values were also flagged as potential outliers that should
be investigated more carefully.

@@

\smallskip Control input variables, looks and meta information\\
\quad \code{useVar} & What variables should be used? & \code{NULL} (corresponding to all variables) \\
\quad \code{ordering} & Ordering of the variables in the data summary (as is or alphabetical) & \code{"asIs"} \\
\quad \code{onlyProblematic} & Should only variables flagged as problematic be included in the \textit{Variable list}? & \code{FALSE} \\
%\quad \code{listChecks} & Should an overview of what checks were performed be listed in the \textit{Data report overview}? &  \code{TRUE} \\
%slettet for at få plads til nye argumenter
\quad \code{preChecks} & What check functions should be called to determine whether a variable is suitable for summarization, visualization and checking? & \code{c("isKey", "isSingular", "isSupported")} \\
\quad \code{reportTitle} & What should the title displayed on the front page of the report be? & \code{NULL} (corresponds to the dataset name) \\
\quad \code{twoCol} & Should the summary table and visualizations be placed side-by-side (in two columns)? & \code{TRUE} \\

\smallskip Control summarize, visualize, and check steps \\
\quad \code{summaries} & What summaries should be performed for each variable type? & See Table~\ref{table.SVCfunctions} \\
\quad \code{visuals} & What type of visualization should be provided for each variable type? & See Table~\ref{table.SVCfunctions} \\
\quad \code{checks} & What checks should be applied to each variable type? & See Table~\ref{table.SVCfunctions} \\
\quad \code{mode} & What steps should be performed for each variable (out
                 of the three possibilities \textit{summarize},
                 \textit{visualize}, \textit{check})? &
                                                        \code{c("summarize", "visualize", "check")}  \\
\quad \code{smartNum} & Should numerical values with only a few unique
                     levels be flagged and treated as a factor variable? & \code{TRUE} \\
\quad \code{maxProbVals} & Maximum number of problematic values to print, if any are found in data checks & \code{10} \\
\quad \code{maxDecimals} & Maximum number of decimals to print for numeric values in the variable list & \code{2} \\
\quad \code{treatXasY} & How should non-supported variable classes be handled? & \code{NULL} (no handling) \\
%Takes a list on the format \code{list(X = Y)} where \code{X} is a non-supported variable class (e.g. \code{complex}) and \code{Y} is a supported one (e.g. \code{numeric}) & \code{NULL} (No rules for handling non-supported variable types) \\

\smallskip Control output and post-processing \\
\quad \code{output} & Type of output file to be produced (html, word (.docx) or pdf) & \code{"pdf"} \hl{OBS. Ændrer vi dette?}\\
\quad \code{render} & Should the output file be rendered from markdown? & \code{TRUE} \\
\quad \code{openResult} & If a  pdf/html file is rendered, should it
                       automatically open afterwards, and if not,
                       should the \code{rmarkdown} file be opened? & \code{TRUE} \\
\quad \code{replace} & Overwrite an existing file with the same name? & \code{FALSE} \\
\quad \code{vol} & Add a suffix to the file name of the outputted report &   \code{""} (no suffix)\\

@@

\subsection{Polishing off the arguments}
We begin with an example that is intended as an illustration of how
\code{makeDataReport()} might be used in the very first stages of data
cleaning, when we are uncertain about the complexities of the
errors and how much time should be allocated to data cleaning. At
this stage, what is really needed, is a very rough idea of the
severity of errors in the dataset. In this scenario, we might wish to
obtain a summary document in html format that only contains the
variables with potential problems, and with a limit of, say, maximum
2 printed potential problematic value per check for each variable. Also, we can add the
argument \code{replace = TRUE} in order to force \code{makeDataReport()} to
overwrite any existing files produced by \code{makeDataReport()}.  Using the
\code{toyData} dataset as a guinea pig, we type:

@@

The final rendering of the generated markdown file is controlled by
the \code{render} and \code{openResult} arguments, which both default to
\code{TRUE}. \code{render} determines if the \proglang{R} markdown file
produced should be rendered using the \pkg{rmarkdown} \citep{rmarkdown} package and
\code{openResult} decides whether the outputted file should be
opened. The following command produces an \proglang{R} markdown file
containing the information needed for generating a data report, but without
rendering nor opening the markdown file:

@@

\begin{table}
\centering
\begin{tabular}{p{0.35\linewidth} p{0.3\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth} p{0.01\linewidth}
 p{0.01\linewidth} p{0.01\linewidth}}
  \hline
& Description &  \multicolumn{7}{c}{Variable classes} \\ \smallskip
 & &  C & F & I & L & B & N & D\\
  \hline \smallskip
  \textbf{\code{summaryFunction}s}  \smallskip \\
  \quad \code{centralValue} & Compute median for numeric variables, mode for categorical variables &  \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} \\
  \quad \code{countMissing} & Compute proportion of missing observations &  \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$}  \\
  \quad \code{minMax} & Find minimum and maximum values &   &  & \blue{$\times$} & &  & \blue{$\times$} & \blue{$\times$}  \\
  \quad \code{quartiles} & Compute 1st and 3rd quartiles &    &  & \blue{$\times$} & &  & \blue{$\times$} & \blue{$\times$} \\
  \quad \code{uniqueValues} & Count number of unique values &   \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$}  \\
  \quad \code{variableType} & Data class of variable & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$}  \\
  \smallskip \\

 \textbf{\code{visualFunction}s} \smallskip \\
  \quad \code{basicVisual} & Histograms and barplots using base \proglang{R} graphics &  $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ \\
  \quad \code{standardVisual} & Histograms and barplots using \pkg{ggplot2} &  \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} \\
  \smallskip \\

 \textbf{\code{checkFunction}s} \smallskip \\
 \quad \code{identifyCaseIssues} & Identify case issues &  \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{identifyLoners} & Identify levels with $<$ 6 obs. & \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{identifyMissing} & Identify miscoded missing values &  \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} &  \\
 \quad \code{identifyNums} & Identify misclassified numeric or integer variables & \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{identifyOutliers} & Identify outliers &  & & \blue{$\times$} & & \blue{$\times$} & \blue{$\times$} \\
 \quad \code{identifyOutliersTBStyle} & Identify outliers (Turkish Boxplot style) &  & & $\times$ & & $\times$ & $\times$ \\
 \quad \code{identifyWhitespace} & Identify prefixed and suffixed whitespace &  \blue{$\times$} & \blue{$\times$} & & \blue{$\times$} & & &  \\
 \quad \code{isCPR} & Identify Danish CPR numbers & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ & $\times$ &$\times$   \\
 \quad \code{isSingular} & Check if the variable contains only a single value & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$}  \\
 \quad \code{isKey} & Check if the variable is a key & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} \smallskip   \\
 \quad \code{isSupported} & Check if the variable is among the supported variable types & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} & \blue{$\times$} \smallskip   \\
 \hline
\end{tabular}
\caption{Overview of all summary-, visual- and  check functions currently implemented in \pkg{dataMaid}. The variable
  classes C, F, I, L, B, N, and D, refer to \code{character}, \code{factor},
  \code{integer}, \code{labelled}, \code{logical} (boolean), \code{numeric}, and \code{Date}, respectively. The default settings of \code{makeDataReport()} are marked in blue.}
\label{table.SVCfunctions}
\end{table}

\subsection{Controlling contents: What summaries, visualizations and checks should be performed?}
We will now move on to discuss how not only the structure of the
data assesment steps is manipulated, but also its very contents. This is
done through the summarize/visualize/check (SVC) steps, as illustrated
in Figure \ref{figure:cleanStructure}.  \pkg{dataMaid} uses three
different types of functions for performing these steps, namely
\code{summaryFunction}s, \code{visualFunction}s and
\code{checkFunction}s.  By default, \code{makeDataReport()} runs selected summary,
visualization and check functions on each variable in the dataset, and
the exact choice of these functions depends on the classes of the
variables. For instance, though detection of outlier values might be
interesting for numerical variables, it holds little meaning for
factor variables, and therefore, numerical and factor variables need
different checks. Table~\ref{table.SVCfunctions} lists all available
summarize/visualize/check functions, but we can also use the
\code{allSummaryFunctions()}, \code{allVisualFunctions()}, and
\code{allCheckFunctions()} functions in \pkg{dataMaid} to print overview
lists in \proglang{R}. For example, the implemented
\code{summaryFunction}s are:

@@

We can control what summaries and checks are applied for each variable type
through the \code{summaries}, \code{visuals} and \code{checks} arguments of \code{makeDataReport()}. Each of these arguments takes a list with one entry for each variable type and a number of function names for each such entry. The easiest way to specify the arguments is by use of the built-in helper functions \code{setSummaries()}, \code{setVisuals()} and \code{setChecks()} that contain the default settings of \code{makeDataReport()} and simple syntaxes for making small alterations of these default settings.  We can inspect the default settings for summaries by calling:

\begin{Schunk}
\begin{Sinput}
R> setSummaries()
\end{Sinput}
\begin{Soutput}
$character
[1] "variableType" "countMissing" "uniqueValues" "centralValue"

$factor
[1] "variableType" "countMissing" "uniqueValues" "centralValue"

$labelled
[1] "variableType" "countMissing" "uniqueValues" "centralValue"

$numeric
[1] "variableType" "countMissing" "uniqueValues" "centralValue" "quartiles"   
[6] "minMax"

$integer
[1] "variableType" "countMissing" "uniqueValues" "centralValue" "quartiles"   
[6] "minMax"

$logical
[1] "variableType" "countMissing" "uniqueValues" "centralValue"
